Discussion Draft

### 2020-03-24 doc/Discussion.rmd
edited outline for reversed HetC section

@@

3. another thing to point out / regaurding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explaination proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromsoome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference woud have chromatin organization with shorter axis compared to the other sex. There is a plefora of data of sexual dimorphism in linkage map based interference strength indeicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to idenification of specific chromosomes between cells.)

@@

```{r Evolve.tratis.outline, include=FALSE}

#it's not the sex differences that distinguish

@@

# while DSBs have evolved between these groups - not integrated into the table for models -- since most models don't integrate this stage of the pathway.

#1.reversed heterochiasmy direction
#faster -- rapid evoluiton in males

#2 interference strength (intro)

#SACE could fit ...

#spindle based selection 
```

- the male specific differences in meiotic traits are a bit trickier to explain using the above models (since they were designed to explain broad scale pattern).

The general patterns observed from the within strain / between sex data --generate predictions for the relationship between chromatin compaction and interference strength. (the predictions are). Based on the general / braod scale patterns, those predictions are not met -- but we argue that our data are not suited / don't have enough power to -- overcome the weaker signal within the nose of chromosome specific and within nucleus correlation in chromatin compaction. (also distingushing the recombination rate evolution at this scale is not the focus of this manuscript.)

Some of the traits related to faster male evolution -- distinguish in which models can redict them: the reverseal of heterochiasmy direction and the correlated evolution of gwRR and interference strength.

This shifts support away from evolution at the CO:NCO decision to an earlier prophase stage such as when the chromatin compaction is set up. **increase in DSBs is driven by the number of DNA loops** and due to the inverse relationship of loop and chromatin axis length, this would result in longer sc lengths (which reflect the underlying chromatin axis length).

# revered heterochiasmy

```{r reverse.HetC, echo=FALSE}

@@

models for haploid selection -- this **reversal of the direction of heterochiasmy** would be accompanied by (vast changes in haploid stages or degree of imprinting)

the meiotic drive -- model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers.

A reverselal in heteroschiasmy is not clearly predicted by the sace model (unless with the -- except under the conditions discribed below)

# evolution in Interference

@@

>>>>

Advice from Bret:

- Don't know where to fit in DSB evolution results

@@

Another thing to point out / regaurding this model -- is the connection to sex difference in interference strength. Our results of conserved within strain raw IFD lengths - yet when the IFDs are controled for total SC length, we observed sexual dimorphism with males having longer IFD^normalized^ (stronger interference), mirror those observed in [@petkov2007].
The model / explaination proposed in that paper is that The chromatin structure which has longer loops (and shorter axis) will have more DNA bps folded / sequestered into the DNA loops. Thus the strength of interference will be stronger because with each step (count) along the 2D axis, more bases pairs are being skipped above the axis. These findings point out two things, This part of heterochiasmy generates testable prediction: First, the interference strength measured in chromsoome axis units (or SC) is generally conserved between sexes. Second the sex with longer axis and shorter loops will have weaker interference compared to the other sex. Or conversely, the sex with stronger linkage map based interference woud have chromatin organization with shorter axis compared to the other sex. There is a plefora of data of sexual dimorphism in linkage map based interference strength indeicating sexual dimorphism, but comparative cytological data is lacking (difference in perparness of some sexes and complications due to idenification of specific chromosomes between cells.)

@@

```{r Evolve.traits.outline, include=FALSE}

@@

#models weren't designed to explain within species variation in recombination rates 
#or evolution

@@

A thing about this study that we were very interested in was examining the faster evolution in males with the high recombining rates.

BUT applying these above models to the traits connected to the faster male evolution is a bit tricker since none of these moedls were designed to explain within species variation in recombination rates. (most were designed to explain the general pattern of sexual dimorphism in meiotic recombination rates.)

# Revered heterochiasmy

```{r reverse.HetC, echo=FALSE}
# transition connection to the hetC models

#our results - total gwRR isn't the most conserved, (we observed 3 other traits more)

#what does a reversal of HetC imply?  gwRR
#driven by faster evolution in males

#our model had faster evolution in males resulting in a reversal of the canonical pattern of heterichiasmy ( F gwRR > M gwRR)
#other examples, cattle, marsupials

#what do models say about faster evolution in one sex? (yes -- but almost always for canonical pattern of heterochiasmy)

Our results demonstrate that total gwRR isn't the most conserved, (we observed 3 other traits more conserved sexual dimorphism examined above). Also our data shovariyng degrees of heterochiasmy magnitude (0 to gough). We the general pattern 2 strains attribute this pattern of reversed heterochiasmy to faster evolution of gwRR in two strains males. Which models predict dfferent rates of evolution between the sexes. Some of the models could explain a reversal of the direction of sexaul dimrohpsim, however most all of the predictions for faster evolution result in males having reduced recombination rates. Under the haploid selection model a  **reversal of the direction of heterochiasmy** would be accompanied by vast changes in haploid stages or degree of imprinting. Following the meiotic drive based model, females should be predicted to have larger linkage maps/gwRR since they are more exposed / weaker to true meiotic drivers. A reversal in heteroschiasmy is not clearly predicted by the SACE model (unless with the -- except under the conditions discribed below). Under the sindle selection model, the male (centrosome/effect SAC) metaphase could drive directional selection (and faster evolution) in males compared to female. However the mechanism/aspect of different fitness is not clear. Under the COM model, evolutionary predictions are not clear (explicit), but following the logic would come from changes in axis length (to generate another wave point) -- which might be due to meiocyte volumne.

# Evolution in Interference

@@

<Some of the traits related to faster male evolution -- distinguish in which models can redict them: the reverseal of heterochiasmy direction and the correlated evolution of gwRR and interference strength.

This shifts support away from evolution at the CO:NCO decision to an earlier prophase stage such as when the chromatin compaction is set up. **increase in DSBs is driven by the number of DNA loops** and due to the inverse relationship of loop and chromatin axis length, this would result in longer sc lengths (which reflect the underlying chromatin axis length).>
